---
title: "Applied Statistical Programming - Spring 2022"
author: Cecilia Sui
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)

# Change eval=FALSE in the code block. Install packages as appropriate.
library(fivethirtyeight)
library(tidyverse)
library(dplyr)
library(readr)
library(tidytext)
library(tm) 
library(lubridate)
library(wordcloud)
library(wordcloud2)
# resolve memory issue 
# library(usethis) 
# usethis::edit_r_environ()
```

\begin{center}
{\Large{\textbf{Problem Set 4}}} \\
\vspace{4 bp}
Due Wednesday, March 16, 10:00 AM (Before Class) \\
\end{center}

\section*{Instructions}
\begin{enumerate}
  \item The following questions should each be answered within an Rmarkdown file. Be sure to provide many comments in your code blocks to facilitate grading. Undocumented code will not be graded.
  \item Work on git. Continue to work in the repository you forked from \url{https://github.com/johnsontr/AppliedStatisticalProgramming2022} and add your code for Problem Set 4. Commit and push frequently. Use meaningful commit messages because these will affect your grade.
  \item You may work in teams, but each student should develop their own Rmarkdown file. To be clear, there should be no copy and paste. Each keystroke in the assignment should be your own.
  \item For students new to programming, this may take a while. Get started.
\end{enumerate}

\section*{\texttt{tidyverse}}



Your task in this problem set is to combine two datasets in order to observe how many endorsements each candidate received using only \texttt{dplyr} functions. Use the same Presidential primary polls that were used for the in class worksheets on February 28 and March 2.

```{r}
# URL to the data that you've used.
# url <- 'https://jmontgomery.github.io/PDS/Datasets/president_primary_polls_feb2020.csv'
polls <- read_csv("president_primary_polls_feb2020.csv")
Endorsements <- endorsements_2020 # from the fiverthirtyeight package
```


First, create two new objects \texttt{polls} and \texttt{Endorsements}. Then complete the following.
\begin{itemize}
  \item Change the \texttt{Endorsements} variable name endorsee to \texttt{candidate\_name}.
  \item Change the \texttt{Endorsements} dataframe into a \texttt{tibble} object.
  \item Filter the \texttt{poll} variable to only include the following 6 candidates: Amy Klobuchar, Bernard Sanders,Elizabeth Warren, Joseph R. Biden Jr., Michael Bloomberg, Pete Buttigieg \textbf{and} subset the dataset to the following five variables: \texttt{candidate\_name, sample\_size, start\_date, party, pct}
  \item Compare the candidate names in the two datasets and find instances where the a candidates name is spelled differently i.e. Bernard vs. Bernie. Using only \texttt{dplyr} functions, make these the same across datasets. 
  \item Now combine the two datasets by candidate name using \texttt{dplyr} (there will only be five candidates after joining).
  \item Create a variable which indicates the number of endorsements for each of the five candidates using \texttt{dplyr}.
  \item Plot the number of endorsement each of the 5 candidates have using \texttt{ggplot()}. Save your plot as an object \texttt{p}.
  \item Rerun the previous line as follows: \texttt{p + theme\_dark()}. Notice how you can still customize your plot without rerunning the plot with new options.
  \item Now, using the knowledge from the last step change the label of the X and Y axes to be more informative, add a title. Save the plot in your forked repository.
\end{itemize}


1. Change the \texttt{Endorsements} variable name endorsee to \texttt{candidate\_name}

```{r}
# rename() to change variable name 
Endorsements <- rename(Endorsements, candidate_name = endorsee)
# check to see whether the change took place  
names(Endorsements)

```


2. Change the \texttt{Endorsements} dataframe into a \texttt{tibble} object.
```{r}
Endorsements <- as_tibble(Endorsements)
# check if the class includes tibble 
class(Endorsements)
```


3. Filter the \texttt{polls} dataset to only include the following 6 candidates: Amy Klobuchar, Bernard Sanders,Elizabeth Warren, Joseph R. Biden Jr., Michael Bloomberg, Pete Buttigieg \textbf{and} subset the dataset to the following five variables: \texttt{candidate\_name, sample\_size, start\_date, party, pct}
```{r}
# create a new dataset to store the changes 
polls_filtered <- polls %>% filter(candidate_name %in% c("Amy Klobuchar",
                                       "Bernard Sanders",
                                       "Elizabeth Warren",
                                       "Joseph R. Biden Jr.",
                                       "Michael Bloomberg",
                                       "Pete Buttigieg")) %>%
                            select(candidate_name,
                                   sample_size,
                                   start_date,
                                   party,
                                   pct)
```

4. Compare the candidate names in the two datasets and find instances where the a candidates name is spelled differently i.e. Bernard vs. Bernie. Using only \texttt{dplyr} functions, make these the same across datasets.
```{r}
# compare the names of candidates of the two datasets 
# identify differences 
sort(unique(polls_filtered$candidate_name))
sort(unique(Endorsements$candidate_name))

# use dplyr to make them the same
polls_filtered <- polls_filtered %>%
  # "Bernard Sanders" --> "Bernie Sanders"
  mutate(candidate_name = replace(candidate_name,
                          candidate_name == "Bernard Sanders",
                          "Bernie Sanders" )) %>%
  # "Joseph R. Biden Jr." --> "Joe Biden"
  mutate(candidate_name = replace(candidate_name,
                          candidate_name == "Joseph R. Biden Jr.",
                          "Joe Biden"))

# Check to see if changes are made 
sort(unique(polls_filtered$candidate_name))
sort(unique(Endorsements$candidate_name))
```


5. Now combine the two datasets by candidate name using \texttt{dplyr} (there will only be five candidates after joining).
```{r}
joined_df <- merge(polls_filtered, Endorsements, 
                       by = "candidate_name")
unique(joined_df$candidate_name)
```

6. Create a variable which indicates the number of endorsements for each of the five candidates using \texttt{dplyr}.
```{r}
cnt_endorsement <- joined_df %>% 
                   group_by(candidate_name) %>%
                   count() %>% 
                   rename(endorsement_count = n)
cnt_endorsement
```


7. Plot the number of endorsement each of the 5 candidates have using \texttt{ggplot()}. Save your plot as an object \texttt{p}.
```{r}
p <- ggplot(cnt_endorsement, aes(x = factor(candidate_name),
                                 y = endorsement_count,
                                 fill = cnt_endorsement$candidate_name,
                                 label = cnt_endorsement$endorsement_count)) +
  geom_bar(stat = "identity", width = 0.5) +
  xlab("Candidate Name") +
  ylab("Number of endorsement") + 
  scale_fill_discrete(name = "Candidates") +
  geom_text(size = 3, position = position_stack(vjust = 0.9))  
p
```

8. Rerun the previous line as follows: \texttt{p + theme\_dark()}. Notice how you can still customize your plot without rerunning the plot with new options.
```{r}
p + theme_dark()
```

9. Now, using the knowledge from the last step change the label of the X and Y axes to be more informative, add a title. Save the plot in your forked repository.
```{r}
p <- p + theme_classic() +
  ggtitle("Candidate Endorsement Comparison") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=0.5))
p

# save to file
# please see github repo
# ggsave("endorsement_plot.pdf", p)
```





\newpage 

\section*{Text-as-Data with \texttt{tidyverse}}

For this question you will be analyzing Tweets from President Trump for various characteristics. Load in the following packages and data:

```{r message = FALSE, warning=FALSE}
# trump_tweets_url <- 'https://politicaldatascience.com/PDS/Datasets/trump_tweets.csv'
tweets <- read_csv("trump_tweets.csv")
```

\begin{itemize}
  \item First separate the \texttt{created\_at} variable into two new variables where the date and the time are in separate columns. After you do that, then report the range of dates that is in this dataset.
  \item Using \texttt{dplyr} subset the data to only include original tweets (remove retweents) and show the text of the President's \textbf{top 5} most popular and most retweeted tweets. (Hint: The \texttt{match} function can help you find the index once you identify the largest values.) 
  \item Create a \textit{corpus} of the tweet content and put this into the object \texttt{Corpus} using the \texttt{tm} (text mining) package. (Hint: Do the assigned readings.)
  \item Remove extraneous whitespace, remove numbers and punctuation, convert everything to lower case and remove 'stop words' that have little substantive meaning (the, a, it).
  \item Now create a \texttt{wordcloud} to visualize the top 50 words the President uses in his tweets. Use only words that occur at least three times. Display the plot with words in random order and use 50 random colors. Save the plot into your forked repository.\item Create a \textit{document term matrix} called \texttt{DTM} that includes the argument \texttt{ control = list(weighting = weightTfIdf)}
  \item Finally, report the 50 words with the the highest tf.idf scores using a lower frequency bound of .8.
\end{itemize}

1. First separate the \texttt{created\_at} variable into two new variables where the date and the time are in separate columns. After you do that, then report the range of dates that is in this dataset.
```{r}
# modify in place 
tweets <- tweets %>% 
    # create a new col date to store the dates
    mutate(date = str_split_fixed(tweets$created_at, " ", 2)[,1]) %>%
    # create a new col time to store the timestamps 
    mutate(time = str_split_fixed(tweets$created_at, " ", 2)[,2])

# convert to date
tweets$date <- as.Date(tweets$date, "%m/%d/%Y")

# find the range of dates: 
# from "2014-01-01" to "2020-02-14"
max(tweets$date)
min(tweets$date) 
```


2. Using \texttt{dplyr} subset the data to only include original tweets (remove retweents) and show the text of the President's \textbf{top 5} most popular and most retweeted tweets. (Hint: The \texttt{match} function can help you find the index once you identify the largest values.) 
```{r tidy = TRUE}
original_tweets <- tweets %>% 
                  # remove retweets 
                  filter(is_retweet == FALSE)

original_tweets %>% 
          # top 5 most popular
          top_n(n = 5, wt = favorite_count) %>%
          arrange(desc(favorite_count)) %>%
          select(text)

original_tweets %>% 
          # top 5 most retweeted
          top_n(n = 5, wt = retweet_count) %>%
          arrange(desc(retweet_count)) %>%
          select(text)
```


3. Create a \textit{corpus} of the tweet content and put this into the object \texttt{Corpus} using the \texttt{tm} (text mining) package. (Hint: Do the assigned readings.)
```{r}
Corpus <- VCorpus(VectorSource(original_tweets$text))
```


4. Remove extraneous whitespace, remove numbers and punctuation, convert everything to lower case and remove 'stop words' that have little substantive meaning (the, a, it).
```{r}
Corpus <- Corpus %>%
  # remove numbers
  tm_map(removeNumbers) %>%
  # remove punctuation 
  tm_map(removePunctuation, ucp = TRUE) %>%
  # remove whitespace 
  tm_map(stripWhitespace)

# convert to lower case
Corpus <- tm_map(Corpus, content_transformer(tolower))

# remove stop words 
Corpus <- tm_map(Corpus, removeWords, stopwords("english"))

# remove urls
Corpus <- tm_map(Corpus, content_transformer(function(x, pattern) gsub("?(f|ht)tp(s?).*", "", x)))
```


5. Now create a \texttt{wordcloud} to visualize the top 50 words the President uses in his tweets. Use only words that occur at least three times. Display the plot with words in random order and use 50 random colors. Save the plot into your forked repository.
```{r}
# generate term document matrix 
tdm <- TermDocumentMatrix(Corpus) 
matrix <- as.matrix(tdm) 
words <- sort(rowSums(matrix), decreasing = T) 
df <- data.frame(word = names(words), freq = words)


# for reproducibility 
set.seed(12345) 
# Extract color info
palette3_info <- brewer.pal.info[brewer.pal.info$category == "qual", ]  
palette3_all <- unlist(mapply(brewer.pal,                     
                              palette3_info$maxcolors,
                              rownames(palette3_info)))
# Sample colors
palette3 <- sample(palette3_all, 50)

# generate wordcloud
wordcloud(words = df$word[1:50], # top 50 words
          freq = df$freq, 
          min.freq = 3, # occur at least 3 times 
          max.words = 50, 
          random.order = TRUE, 
          rot.per = 0.35,
          colors = palette3,
          scale = c(3,1))
```


6. Create a \textit{document term matrix} called \texttt{DTM} that includes the argument \texttt{ control = list(weighting = weightTfIdf)}
```{r warning = FALSE}
# Create document term matrix 
# Doc as rows and Term as cols 
DTM <- DocumentTermMatrix(Corpus, 
                          control = list(weighting = weightTfIdf))
# inspect(DTM)
```


7. Finally, report the 50 words with the the highest tf.idf scores using a lower frequency bound of .8.
```{r}
# findMostFreqTerms(DTM)
# DTM <- removeSparseTerms(DTM, .8)
df_DTM <- tidy(DTM)
df_DTM %>% 
  # frequency bound of 0.8
  filter(count > 0.8) %>%
  # get top 50 scores 
  slice_max(count , n = 50)
```









